---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The data have been provided in `activity.zip`. Set the working directory to 
the location of the data file. 

```{r, echo=TRUE}
setwd("~/Documents/DataScienceJHU/RWork/reproducible/RepData_PeerAssessment1/")
df = read.csv(unz("activity.zip","activity.csv"))
```

## What is mean total number of steps taken per day?
```{r, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
```
```{r, echo=TRUE}
totdf <- summarize(group_by(df,date), 
                   tot=sum(steps)  )
library(ggplot2)
ggplot(data=totdf, aes(totdf$tot)) + 
    geom_histogram(breaks=seq(0, 23000, by = 500), 
                 col="red", 
                 fill="green", 
                 alpha = .2) + 
    labs(x="Total Steps per Day", y="Days") 
```

What are the mean and median number of steps per day?
```{r, echo=TRUE}
mymean = mean(totdf$tot, na.rm=TRUE)
mymedian = median(totdf$tot, na.rm=TRUE)
```
The mean number of steps per day is `r format(mymean,nsmall=2)` and the median steps per day is `r mymedian`.

## What is the average daily activity pattern?
```{r, echo=TRUE}
avedf <- summarize(group_by(df,interval), 
                   ave=mean(steps, na.rm=TRUE)  )
ggplot(data=avedf, aes(interval, ave))  + geom_line() +
    xlab("Interval") + ylab("Average steps in interval")
```

Which 5-minute interval, averaged over all days in the dataset, contains
the maximum number of steps?
```{r, echo=TRUE}
max_steps_interval = max(avedf$ave)
max_interval = avedf[avedf$ave == max(avedf$ave),]$interval
```
On average, the most steps are taken in the interval `r max_interval`.

## Imputing missing values

How many missing values are in the dataset?
```{r, echo=TRUE}
tot_missing = sum(is.na(df$steps))
tot_rows = nrow(df)
```
The total number of missing values is `r tot_missing`, out of a total of `r tot_rows` observations.

One strategy for filling in missing values is to use the mean for the given interval.
```{r, echo=TRUE}
impsteps <- ifelse(is.na(df$steps) == TRUE, 
                  avedf$ave[avedf$interval %in% df$interval], 
                  df$steps)
imputed <- mutate(df, steps = impsteps)
```

```{r, echo=TRUE}
totimp <- summarize(group_by(imputed,date), 
                   tot=sum(steps)  )
ggplot(data=totimp, aes(tot)) + 
    geom_histogram(breaks=seq(0, 23000, by = 500), 
                 col="red", 
                 fill="green", 
                 alpha = .2) + 
    labs(x="Total Steps per Day", y="Days") 
```

What are the mean and median number of steps per day?
```{r, echo=TRUE}
newmean = mean(totimp$tot)
newmedian = median(totimp$tot)
```

The mean number of steps per day from the imputed dataset is `r format(newmean,nsmall=2)` and the median
is `r format(newmedian,nsmall=2)`. For comparison, ignoring NAs in the original data, the mean was
`r format(mymean,nsmall=2)` and the median was `r mymedian`. Imputing the values from the averages
does not make big differences in these numbers, and the histogram is affected only in bins near the mean/median. 

## Are there differences in activity patterns between weekdays and weekends?

Make a factor variable with values of "weekend" or "weekday".
```{r, echo=TRUE}
isweekday <- ifelse((weekdays(as.Date(imputed$date)) == "Sunday") |
                         (weekdays(as.Date(imputed$date)) == "Saturday"), 
                  "weekend", 
                  "weekday")
weekdaydf <- mutate(imputed, weekday = factor(isweekday))
```

```{r, echo=TRUE}
aveimputed <- summarize(group_by(weekdaydf,weekday, interval), 
                   ave=mean(steps)  )
ggplot(aveimputed, aes(interval, ave)) + geom_line() +
    facet_grid(weekday ~ .) +
    labs(x="Interval", y="Average steps in interval")
```

From the plot we can see that weekdays have a larger peak in the 835
interval, presumably as people are walking to work or school. Also,
people are walking earlier in the day on weekdays than on weekends.
